---
title: "Example QQ Plot"
author: "Joanna K"
date: "2024-01-05"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r lib, include=T}
suppressWarnings(suppressMessages(library(tidyverse)))
suppressWarnings(suppressMessages(library(detectnorm)))
theme_set(theme_bw())
```

## Introduction
A normal probability (Q-Q) plot is a probability plot that we use to check whether the variable follows a normal distribution.
If the variable is normally distributed, then if we plot it against their expected normal Z-scores, we will get a diagonal line. --> a standard uniform distribution

## Generating the variables

```{r generating variables}

# normally distributed variables

set.seed(10)

# let's say we have 1000 observations of a variable
n <- 1000
# and the mean value (mi) is 10
mi <- 10
# and the standard deviation (sigma) is 5
sigma <- 5

df <- tibble(
  normal_distr_1 = rnorm(n, mi, sigma)
)

glimpse(df)

x <- df$normal_distr_1

hist(x, 
     xlim = c(min(x), max(x)), 
     probability = TRUE, 
     nclass = max(x) - min(x) + 1, 
     col = 'lightblue')
lines(density(x, bw=1), col = 'blue', lwd = 3)

# Of course these are not completely perfectly normally distributed values.
# If we want perfection, there is a density function in r - dnorm that gives you the normal distribution density based on vector of quantiles.

sample.range <- range(df$normal_distr_1)
values.range <- sample.range[1]:sample.range[2]
normal_density_1 = dnorm(values.range, mi, sigma) 

# this is how the perfect normal distribution would look like  
ggplot() +
geom_point(aes(values.range, normal_density_1))


```
## Adding the probabilities and positions and drawing the QQ Plots

```{r prob and pos}

# 1) Sort the data in ascending or descending order.

df <- df %>% 
  arrange(normal_distr_1)

# 2) Add the positions
# Find the z-value (cut-off point) for each segment.
# this point is observation minus the mean, divided by sigma.
# or you may use the function qnorm.

result = vector("numeric", n) 

for(i in 1:n) {
   x = i/(n+1)
 result[i] <- x
  }
result

df$theoretical_quantiles_1 <- result

df$sample_quantiles_1 <- (df$normal_distr_1 - mean(df$normal_distr_1)) / sd(df$normal_distr_1)

df$probabilities_1 <- pnorm(df$normal_distr_1, mi, sigma)

glimpse(df)
# 3) Draw the qqplots

# theoretical and sample quantiles comparison
df %>% 
  ggplot() +
  geom_point(aes(theoretical_quantiles_1, sample_quantiles_1), alpha = 0.5, color = "steelblue")

# theoretical quantiles and probabilities
df %>% 
  ggplot() +
  geom_point(aes(theoretical_quantiles_1, probabilities_1), alpha = 0.5, color = "steelblue")

# qq plot using the built-in function in ggplot
df %>% 
  ggplot() +
  geom_qq(aes(sample = normal_distr_1), alpha = 0.5, color = "pink") +
  geom_qq_line(aes(sample = normal_distr_1))

# and in base r
qqnorm(df$normal_distr_1, pch = 1, frame = FALSE)
qqline(df$normal_distr_1, col = "steelblue", lwd = 2)

```

Notes: I do not fully understand how the build-in qqplot functions work - the units are different than when I do the plot manually.
I guess there are different ways of drawing the QQ plots.
It's also possible that I made a mistake in calculating the theoretical quantiles - I am still not sure whether I did it correctly.

## Other distributions

```{r other distr}
# try also other distributions?

df <- df %>% 
  mutate(
    skewed_distr_2 = rnonnorm(n = n, mean = mi, sd = sigma, skew = 20, kurt = 0)$dat,
    skewed_distr_3 = rnonnorm(n = n, mean = mi, sd = sigma, skew = 0, kurt = 20)$dat
    
  )
glimpse(df)

x <- df$skewed_distr_2

hist(x, 
     xlim = c(min(x), max(x)), 
     probability = TRUE, 
     nclass = max(x) - min(x) + 1, 
     col = 'lightblue')
lines(density(x, bw=1), col = 'blue', lwd = 3)

x <- df$skewed_distr_3

hist(x, 
     xlim = c(min(x), max(x)), 
     probability = TRUE, 
     nclass = max(x) - min(x) + 1, 
     col = 'lightblue')
lines(density(x, bw=1), col = 'blue', lwd = 3)

# qq plots of these two
# this time let's just do the build-in function plots.

df %>% 
  ggplot() +
  geom_qq(aes(sample = skewed_distr_2), alpha = 0.5, color = "pink") +
  geom_qq_line(aes(sample = skewed_distr_2))

df %>% 
  ggplot() +
  geom_qq(aes(sample = skewed_distr_3), alpha = 0.5, color = "pink") +
  geom_qq_line(aes(sample = skewed_distr_3))

```

## Use the linear regression equation to calculate the alpha and beta hat
We can use linear regression to calculate the function that will draw for us the diagonal line of the QQ plot.
x - positions = theoretical quantiles; 
y - probabilities; 
lm = x ~ y

```{r lin regr}

m <- lm(theoretical_quantiles_1 ~ probabilities_1, data = df)
summary(m)

# calculate the alpha hat, which is the intercept, and beta hat, which is the coefficient

(alpha_hat <- m$coefficients[1])
(beta_hat <- m$coefficients[2])

# adding the function of the theoretical quantiles to the graph
df %>% ggplot() +
  geom_point(aes(theoretical_quantiles_1, probabilities_1), color = "steelblue", alpha = 0.2) +
  geom_function(fun = ~alpha_hat + beta_hat * .x, color = "red")

```

